### 2021-01-23 paper/intro.tex
Egits

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order of reads and writes,
where each read has a matching \emph{fulfilling} write,
for example:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2\SEMI
    r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx1}{\DR{x}{1}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=20,in=160]{wx1}{rx1}
        \wk{wx0}{wx1}
        \wk{wx1}{wy2}
        \wk{wy2}{ry2}
        \wk{ry2}{rx1}
      \end{tikzinline}}
  \end{gather*}
This model naturally extends to the case of shared-memory concurrency
in a natural way, leading to a \emph{sequentially consistent}
semantics, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
    \PAR
    \THREAD{r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx1}{\DR{x}{1}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=20,in=160]{wx1}{rx1}
        \wk{wx0}{wx1}
        \wk{wx1}{wy2}
        \wk{ry2}{rx1}
      \end{tikzinline}}
  \end{gather*}
Unfortunately, this model does not compile efficiently to commodity
hardware, resulting in a X\% increase in CPU time~\cite{???}, and
hence power consumption.  Developers of software and compilers have
therefore been faced with a difficult trade-off, between an elegant
model of memory, and its impact on resource usage (such as size of
data centres, electricity bills and carbon footprint). Unsurprisingly,
many have chosen to prioritize efficiency over elegance.

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location} and not globally
over the whole progfram. This allows executions which
are inconsistent with program order, such as:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
    \PAR
    \THREAD{r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx0}{\DR{x}{0}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=15,in=165]{wx0}{rx0}
        \wk{wx0}{wx1}
        \wk[out=200,in=340]{rx0}{wx1}
      \end{tikzinline}}
  \end{gather*}
In such models, the causal order between events is important,
and includes control and data dependencies, to avoid
paradoxical ``out of thin air'' examples such as:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \FI}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI y\GETS1}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
Unfortunately, while a simple syntactic approach to dependency calculation
suffices for hardware models, it is not preserved by common compiler
optimizarions. For example, if we calculate control dependencies syntactically,
then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
`  \begin{gather*}
    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \ELSE y\GETS1 \FI}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
An optimizing compiler might lift the assignment $y\GETS1$ out of the conditional,
thus removing the control dependency.

@@

\item \emph{syntactic} methods used in hardware models
  such as ARM~\cite{???} or x86-TSO~\cite{???},
\item \emph{alternate universe} methods (which give a semantics based on multiple speculative executions
  of the same program) such as the Java Memory Model~\cite{???},
  the speculative operational semantics of~\cite{???},
  the promising semantics of~\cite{???},
  or the event structures semantics of~\cite{???},
\item \emph{rewriting} methods, which give an operational model
  up to syntactic rewrites, such as~\cite{???}, and
\item \emph{logical} methods, such as the pomsets with preconditions
  model of~\cite{???}.

\end{itemize}
In this paper, we will focus on logical models, as those are compositional,
and align well with existing models of sequential composition.
The heart of the model of~\cite{???} is to add logical preconditions
to events, which are introduced by store (modelling data dependencies)
or conditionals (modelling control dependencies):
  \begin{gather*}
    \IF{s{<}1} \THEN z\GETS r\&s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wz0}{(s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{}
      \end{tikzinline}}
  \end{gather*}
and are discharged by being ordered with a read event:
  \begin{gather*}
    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{right=of ry0}
        \po{ry0}{wz0}
      \end{tikzinline}}
  \end{gather*}
Note that there is dependency order from $(\DR{y}{0})$ to $(\DW{z}{0})$
so the precondition for $(\DW{z}{0})$ only has to be satisfied assuming the hypothesis
$(s=0)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
which is why we do not assume the hypothesis $(r=0)$. The precondition on
$(\DW{z}{0})$ is a tautology, and so can be elided:
  \begin{gather*}
    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \event{wz0}{\DW{z}{0}}{right=of ry0}
        \po{ry0}{wz0}
      \end{tikzinline}}
  \end{gather*}
While existing models of relaxed memory have detailed treatmenmts of parallel composition,
they pften give sequential composition little attention, either ignoring it altogether,
or treating it operationally with its usual small-step semantics. This paper
investigates how existing models of sequential composition interact with relaxed memory.

>>>>

For single-threaded programs, memory can be thought of as you might
expect: programs write to, and read from, memory references.
This can be thought of as a total order of reads and writes,
where each read has a matching \emph{fulfilling} write,
for example:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2\SEMI
    r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx1}{\DR{x}{1}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=20,in=160]{wx1}{rx1}
        \wk{wx0}{wx1}
        \wk{wx1}{wy2}
        \wk{wy2}{ry2}
        \wk{ry2}{rx1}
      \end{tikzinline}}
  \end{gather*}
This model naturally extends to the case of shared-memory concurrency
in a natural way, leading to a \emph{sequentially consistent}
semantics, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
    \PAR
    \THREAD{r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx1}{\DR{x}{1}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=20,in=160]{wx1}{rx1}
        \wk{wx0}{wx1}
        \wk{wx1}{wy2}
        \wk{ry2}{rx1}
      \end{tikzinline}}
  \end{gather*}
Unfortunately, this model does not compile efficiently to commodity
hardware, resulting in a 37--73\% increase in CPU time~\cite{Liu:2019:ASC:3314221.3314611} on ARM, and
hence power consumption.  Developers of software and compilers have
therefore been faced with a difficult trade-off, between an elegant
model of memory, and its impact on resource usage (such as size of
data centers, electricity bills and carbon footprint). Unsurprisingly,
many have chosen to prioritize efficiency over elegance.

This has led to \emph{relaxed memory models}, in which the requirement of
sequential consistency is weakened to only apply \emph{per-location} and not globally
over the whole program. This allows executions which
are inconsistent with program order, such as:
  \begin{gather*}
    \THREAD{x\GETS0\SEMI x\GETS1\SEMI y\GETS2}
    \PAR
    \THREAD{r\GETS y\SEMI s\GETS x}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wx0}{\DW{x}{0}}{}
        \event{wx1}{\DW{x}{1}}{right=of wx0}
        \event{wy2}{\DW{y}{2}}{right=of wx1}
        \event{ry2}{\DR{y}{2}}{right=of wy2}
        \event{rx0}{\DR{x}{0}}{right=of ry2}
        \rf[out=20,in=160]{wy2}{ry2}
        \rf[out=15,in=165]{wx0}{rx0}
        \wk{wx0}{wx1}
        \wk[out=200,in=340]{rx0}{wx1}
      \end{tikzinline}}
  \end{gather*}
In such models, the causal order between events is important,
and includes control and data dependencies, to avoid
paradoxical ``out of thin air'' examples such as:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \FI}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
This candidate execution forms a cycle in causal order, so is disallowed,
but this depends crucially on the control dependency
from $(\DR{x}{1})$ to $(\DW{y}{1})$, and the data dependency
from $(\DR{y}{1})$ to $(\DW{x}{1})$. If either is missing, then this execution
is acyclic and hence allowed. For example dropping the control dependency
results in:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI y\GETS1}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
Unfortunately, while a simple syntactic approach to dependency calculation
suffices for hardware models, it is not preserved by common compiler
optimizations. For example, if we calculate control dependencies syntactically,
then there is a dependency from $(\DR{x}{1})$ to $(\DW{y}{1})$, and therefore a cycle in, the candidate execution:
  \begin{gather*}
    \THREAD{r\GETS x\SEMI \IF r \THEN y\GETS1 \ELSE y\GETS1 \FI}
    \PAR
    \THREAD{s\GETS y\SEMI x\GETS s}
    \\[-.4ex]
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx1}{\DR{x}{1}}{}
        \event{wy1}{\DW{y}{1}}{right=of rx1}
        \event{ry1}{\DR{y}{1}}{right=of wy1}
        \event{wx1}{\DW{x}{1}}{right=of ry1}
        \rf[out=20,in=160]{wy1}{ry1}
        \rf[out=160,in=20]{wx1}{rx1}
        \po{rx1}{wy1}
        \po{ry1}{wx1}
      \end{tikzinline}}
  \end{gather*}
An optimizing compiler might lift the assignment $y\GETS1$ out of the conditional,
thus removing the control dependency.

@@

\item \emph{syntactic} methods used in hardware models
  such as ARM or x86-TSO \cite{alglave},
\item \emph{alternate universe} methods (which give a semantics based on multiple speculative executions
  of the same program) such as the Java Memory Model~\cite{Manson:2005:JMM:1047659.1040336} 
  and related models \cite{Jagadeesan:2010:GOS:2175486.2175503, DBLP:conf/popl/KangHLVD17, DBLP:journals/pacmpl/ChakrabortyV19},
  % the speculative operational semantics of~\cite{Jagadeesan:2010:GOS:2175486.2175503},
  % the promising semantics of~\cite{DBLP:conf/popl/KangHLVD17},
  % or the event structures semantics of~\cite{DBLP:journals/pacmpl/ChakrabortyV19},
\item \emph{rewriting} methods, which give an operational model
  up to syntactic rewrites, such as~\cite{Pichon-Pharabod:2016:CSR:2837614.2837616}, and
\item \emph{logical} methods, such as the pomsets with preconditions
  model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20}.

\end{itemize}
In this paper, we will focus on logical models, as those are compositional,
and align well with existing models of sequential composition.
The heart of the model of~\cite{DBLP:journals/pacmpl/JagadeesanJR20} is to add logical preconditions
to events, which are introduced by store actions (modeling data dependencies)
and conditionals (modeling control dependencies):
  \begin{gather*}
    \IF{s{<}1} \THEN z\GETS r\&s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{wz0}{(s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{}
      \end{tikzinline}}
  \end{gather*}
Preconditions are discharged by being ordered after a read:
  \begin{gather*}
    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \event{wz0}{(s{=}0) \limplies (s {<} 1) \land (r\&s){=}0 \mid \DW{z}{0}}{right=of ry0}
        \po{ry0}{wz0}
      \end{tikzinline}}
  \end{gather*}
Note that there is dependency order from $(\DR{y}{0})$ to $(\DW{z}{0})$
so the precondition for $(\DW{z}{0})$ only has to be satisfied assuming the hypothesis
$(s{=}0)$. There is no matching order from $(\DR{x}{0})$ to $(\DW{z}{0})$
which is why we do not assume the hypothesis $(r{=}0)$. The precondition on
$(\DW{z}{0})$ is a tautology, and so can be elided in the diagram:
  \begin{gather*}
    r\GETS x\SEMI s\GETS y\SEMI \IF{s{<}1} \THEN z\GETS r\&s \FI
    \\
    \nonumber
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx0}{\DR{x}{0}}{}
        \event{ry0}{\DR{y}{0}}{right=of rx0}
        \event{wz0}{\DW{z}{0}}{right=of ry0}
        \po{ry0}{wz0}
      \end{tikzinline}}
  \end{gather*}
While existing models of relaxed memory have detailed treatmenmts of parallel composition,
they often give sequential composition little attention, either ignoring it altogether,
or treating it operationally with its usual small-step semantics. This paper
investigates how existing models of sequential composition interact with relaxed memory.
